---
title: "R for Data Science Project Statistical Analyses"
author: "Benjamin Poag"
date: '2022-04-14'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Import libraries
library(tidyverse)
library(readxl) # For reading in an Excel sheet
library(naniar) # For replacing values with NA
library(psych) # For finding significance levels for correlations
library(stargazer) # For well-formatted tables
library(broom) # For informative ANOVA tables
library(rpart) # For classification trees
library(summarytools) # For frequency tables
library(xtable) # For formatted correlation matrix
library(ggplot2) # For flexible plotting
library(rattle) # For formatting classification trees
library(rpart.plot)
library(RColorBrewer)
library(nnet) # For multinomial regression analysis

# Import data 
pwe <- read.csv("PWE_data_formatted.csv")
```

## Clean data
```{r}
# Remove unnecessary variables
pwe <- pwe[ , !(names(pwe) %in% c(paste0("Q", 1:19, "I"), paste0("Q", 1:19, "E"), paste0("VCL", 1:16)))]

# Recode reverse-keyed items
pwe[ , c("Q9A", "Q13A", "Q15A")] <- 6 - pwe[ , c("Q9A", "Q13A", "Q15A")]
pwe[ , c("TIPI2", "TIPI4", "TIPI6", "TIPI8", "TIPI10")] <- 8 - pwe[ , c("TIPI2", "TIPI4", "TIPI6", "TIPI8", "TIPI10")]

# Find totals for each key variable
pwe <- pwe %>% mutate(pwe = select(., paste0("Q", 1:19, "A")) %>% rowSums()) %>% mutate(extraversion = select(., c("TIPI1", "TIPI6")) %>% rowSums()) %>% mutate(agreeableness = select(., c("TIPI2", "TIPI7")) %>% rowSums()) %>% mutate(conscientiousness = select(., c("TIPI3", "TIPI8")) %>% rowSums()) %>% mutate(emotional_stability = select(., c("TIPI4", "TIPI9")) %>% rowSums()) %>% mutate(openness = select(., c("TIPI5", "TIPI10")) %>% rowSums())

```

## Replace important numbered responses with character strings
```{r}
# Replace education level labels
pwe["education"][pwe["education"] == 0] <- NA
pwe["education"][pwe["education"] == 1] <- "Less than high school"
pwe["education"][pwe["education"] == 2] <- "High school"
pwe["education"][pwe["education"] == 3] <- "University degree"
pwe["education"][pwe["education"] == 4] <- "Graduate degree"

# Replace gender labels
pwe["gender"][pwe["gender"] == 0] <- NA
pwe["gender"][pwe["gender"] == 1] <- "Male"
pwe["gender"][pwe["gender"] == 2] <- "Female"
pwe["gender"][pwe["gender"] == 3] <- "Other"

# Replace religion labels
pwe["religion"][pwe["religion"] == 0] <- NA
pwe["religion"][pwe["religion"] == 1] <- "Agnostic"
pwe["religion"][pwe["religion"] == 2] <- "Atheist"
pwe["religion"][pwe["religion"] == 3] <- "Buddhist"
pwe["religion"][pwe["religion"] == 4] <- "Christian (Catholic)"
pwe["religion"][pwe["religion"] == 5] <- "Christian (Mormon)"
pwe["religion"][pwe["religion"] == 6] <- "Christian (Protestant)"
pwe["religion"][pwe["religion"] == 7] <- "Christian (Other)"
pwe["religion"][pwe["religion"] == 8] <- "Hindu"
pwe["religion"][pwe["religion"] == 9] <- "Jewish"
pwe["religion"][pwe["religion"] == 10] <- "Muslim"
pwe["religion"][pwe["religion"] == 11] <- "Sikh"
pwe["religion"][pwe["religion"] == 12] <- "Other"

# Replace race labels
pwe["race"][pwe["race"] == 10] <- NA
pwe["race"][pwe["race"] == 11] <- "Asian"
pwe["race"][pwe["race"] == 12] <- "Arab"
pwe["race"][pwe["race"] == 13] <- "Black"
pwe["race"][pwe["race"] == 14] <- "Indigenous Australian"
pwe["race"][pwe["race"] == 15] <- "Native American"
pwe["race"][pwe["race"] == 16] <- "White"
pwe["race"][pwe["race"] == 17] <- "Other"




```

## Obtain descriptive statistics
```{r}
# Structure of dataset
str(pwe)

# Summarize data
summary(pwe)

# Determine percentage of missing data
total.NA <- sum(is.na(pwe))
total.cells <- prod(dim(pwe))
perc.missing <- total.NA / total.cells * 100

# Descriptive statistics for demographic variables

## Calculate frequency of each demographic category
demo.vars <- subset(pwe, select = c(education, gender, religion, race))
summary(demo.vars)
demo.table <- freq(demo.vars)
demo.table

# Calculate and tabulate means and standard deviations for continuous variables
total.vals <- subset(pwe, select = c(pwe, extraversion, agreeableness, conscientiousness, emotional_stability, openness, age, familysize))
stargazer(total.vals, type = 'text', summary.stat = c("n", "mean", "sd"), digits=2, align=T, title = "Descriptive Statistics", out="desc_table.htm")

# Calculate and tabulate correlation matrix plus significance levels
pwe.corr <- round(cor(total.vals), 2)
stargazer(pwe.corr, title = "Correlation Matrix", out = "corr_mat.htm")
pwe.corr.test <- corr.test(total.vals)$p
pwe.corr.test
```

## Calculate Cronbach's alphas for measures used
```{r}
library(ltm) # For calculating scale alphas

# Calculate alphas for measures in data
pwe_alpha <- cronbach.alpha(pwe[paste0("Q", 1:19, "A")])
e_alpha <- cronbach.alpha(pwe[c("TIPI1", "TIPI6")])
a_alpha <- cronbach.alpha(pwe[c("TIPI2", "TIPI7")])
c_alpha <- cronbach.alpha(pwe[c("TIPI3", "TIPI8")])
es_alpha <- cronbach.alpha(pwe[c("TIPI4", "TIPI9")])
o_alpha <- cronbach.alpha(pwe[c("TIPI5", "TIPI10")])
```

## Multiple regression analysis for personality traits
```{r}
# Create multiple regression model for PWE with personality traits as predictors
mult.reg <- lm(pwe ~ extraversion + agreeableness + conscientiousness + emotional_stability + openness + age + familysize, data = total.vals)
summary(mult.reg)$coefficient
stargazer(mult.reg,
          title="Table 3. Multiple Regression of PWE on Personality",
          dep.var.caption="DV: Protestant Work Ethic Scale Score",
          dep.var.labels = c("PWE"),
          covariate.labels=c("Extraversion", "Agreeableness", "Conscientiousness", "Emotional Stability", "Openness", "Age", "Family Size"),
          notes.label="Significance levels",
          type="html",
          out="mult_reg.htm")
```

## Two-way ANOVA for race and religious beliefs
```{r}
# Conduct two-way ANOVA for factors of race and religious beliefs plus their interaction
pwe.aov <- aov(pwe ~ as.factor(race) * as.factor(religion), data = pwe)

# Build table of ANOVA output
summary.aov(pwe.aov)
tidy_aov <- tidy(pwe.aov)

# Conduct Tukey's post hoc test
tidy_Tukey <- tidy(TukeyHSD(pwe.aov, which = c("as.factor(race)", "as.factor(religion)")))
tidy_Tukey <- tidy_Tukey %>% filter(adj.p.value < 0.05)
```

## Visualize frequency of education levels
```{r}
# Use ggplot to visualize race
ggplot(data = pwe) +
  geom_bar(aes(race, fill = race))  +
  labs(fill = "Race") +
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1))

# Use ggplot to visualize religious beliefs
ggplot(data = pwe) +
  geom_bar(aes(religion, fill = religion))  +
  labs(fill = "Religion") +
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1))

# Use ggplot to visualize education level
ggplot(data = pwe) +
  geom_bar(aes(education, fill = education))  +
  labs(fill = "Education Level") +
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1))
```

## Test significance of PWE as predictor of educational level with multinomial logistic regression
```{r}
# Build regression model for education level
edu.regr <- multinom(education ~ pwe, data = pwe)
summary(edu.regr)

# Calculate p value for model
z <- summary(edu.regr)$coefficients/summary(edu.regr)$standard.errors
p.edu <- (1 - pnorm(abs(z), 0, 1)) * 2
p.edu

# Export table of regression statistics
stargazer(edu.regr,
          title="Multinomial Logistic Regression of Education Level on PWE",
          single.row = TRUE,
          column.sep.width = "2pt",
          dep.var.caption="DV: Education Level",
          covariate.labels=c("PWE", "(Intercept)"),
          notes.label="Significance levels",
          type="html",
          out="edu_log_reg.htm")

```

## Construct education classification tree using rpart library
```{r}
# Grow tree with education data
edu.fit <- rpart(education ~ pwe,
   method = "class", data = pwe)

printcp(edu.fit) # display the results
plotcp(edu.fit) # visualize cross-validation results
summary(edu.fit) # detailed summary of splits

# Plot tree
plot(edu.fit, uniform = TRUE,
   main="Classification Tree for Education Level")
text(edu.fit, use.n = TRUE, all = TRUE, cex = .8)

# Create fancy plot of tree
fancyRpartPlot(edu.fit, caption = "Regression Tree for Education Level")
```

## Prune education classification tree 
```{r}
# Prune the education classification tree
p.edu.fit<- prune(edu.fit, cp = edu.fit$cptable[which.min(edu.fit$cptable[,"xerror"]),"CP"]) # cp from cptable   

# plot the pruned tree
plot(p.edu.fit, uniform=TRUE,
   main="Pruned Regression Tree for Education Level")
text(p.edu.fit, use.n=TRUE, all=TRUE, cex=.8)

# Create fancy plot of tree
fancyRpartPlot(p.edu.fit, caption = "Pruned Regression Tree for Education Level")
```